---
title: "EDA Code Notebook"
author: "Kimberly Cardinale, Noelle Crichton, and William Kong"
output: html_document
date: "2025-07-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction:
This dataset contains monthly price data for a wide array of global commodities and price indices from January 1980 to June 2016. The data is sourced from the International Monetary Fund’s Primary Commodity Price System and compiled by the dataset’s [GitHub repository](https://github.com/datasets/commodity-prices/tree/main).

Given the significant impact of energy resources on the world economy, this analysis focuses on three major energy commodities: Crude Oil, Natural Gas, and Coal. Analyzing the historical price movements and relationships between these commodities can provide valuable insights into global economic trends, the impacts of energy policies, and how fossil fuels are economically linked.

This exploratory data analysis (EDA) seeks to:  

1. Explore the price distributions for each commodity.  

2. Visualize long-term trends in prices.  

3. Quantify correlations between the commodities.

This process aims to lay the foundations for exploring research questions such as:  

1. How closely are energy commodities correlated?  

2. What global events might explain sharp price fluctuations?


## Loading Packages and Commodity Data
```{r}
library(tidyverse)
library(data.table)
library(lubridate)
library(ggcorrplot)

# Load the dataset
data <- read_csv("https://raw.githubusercontent.com/datasets/commodity-prices/refs/heads/main/data/commodity-prices.csv")
```

```{r load}
data<-read.csv("https://raw.githubusercontent.com/datasets/commodity-prices/refs/heads/main/data/commodity-prices.csv")
```


## Data Wrangling: Date Formatting, Filtering, Renaming, and Missing Value Check
This step ensures the dataset is clean and focused on the relevant commodities. We begin by checking the column names and then convert the Date column into proper date format using ymd() from lubridate.

We filter the dataset to include only observations from 1992 to 2017, since earlier years contain missing values for Natural Gas. From the 64-column dataset, we isolate three key variables: Crude Oil, Natural Gas (stored in Fuel.Energy.Index), and Coal. The selected columns are renamed for easier reference, and we confirm the dataset’s structure using summary() and head().

```{r sum}
# Check column names
colnames(data)

# Convert Date to proper format
data <- data %>%
  mutate(Date = ymd(Date))

# Subset only energy-related commodities
energy_data <- data %>%
  filter(Date >= ymd("1992-01-01") & Date <= ymd("2017-06-01")) %>%
  select(Date, `Crude.Oil.petroleum`, `Fuel.Energy.Index`, Coal)

# Rename columns for easier reference
colnames(energy_data) <- c("Date", "CrudeOil", "NaturalGas", "Coal")

# Check for missing values
summary(energy_data)

# Display first 6 rows of cleaned dataset
head(energy_data)
```

The original, raw dataset consisted of 450 rows and 64 columns. Due to missing values in the `NaturalGas` variable from 1980 to 1991, we filtered the data by year to include only the period from 1992 through 2017. The cleaned dataset contains 306 rows and 4 columns, making it ready for exploratory data analysis and visualization.


## Summary Statistics
To understand the central tendencies and spread of commodity prices, we compute summary statistics (mean, median, standard deviation, min, and max) for each commodity. This gives us a baseline understanding of each variable's behavior over time.

```{r}
# Summary statistics
energy_data %>%
  select(-Date) %>%
  summarise_all(list(
    mean = ~mean(., na.rm = TRUE),
    median = ~median(., na.rm = TRUE),
    sd = ~sd(., na.rm = TRUE),
    min = ~min(., na.rm = TRUE),
    max = ~max(., na.rm = TRUE)
  ))
```


## Transforming Data to Long Format for Plotting
To facilitate visualization using ggplot2, we reshape the data from wide to long format. This step is important for plotting multiple commodities in a single figure using facet_wrap() or color grouping.

```{r reshape}
# Reshape to long format for visualization
energy_long <- energy_data %>%
  pivot_longer(cols = -Date, names_to = "Commodity", values_to = "Price")
```


## Visualizing Price Distribution and Median Values by Commodity
We first visualize the distribution of prices for each commodity using histograms. These plots reveal the skewness and spread of each variable and help identify any outliers or unusual patterns. We follow this with boxplots that show the range and median prices for each commodity. Adding median labels provides a quick numerical reference for comparison.

```{r vis}
# Histogram of each commodity price
ggplot(energy_long, aes(x = Price)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 40) +
  facet_wrap(~Commodity, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Energy Commodity Prices (1992–2016)")
```
The histograms show that all three energy commodities have right-skewed price distributions, with most values falling below $100. Coal has the narrowest range, with prices primarily under $50, suggesting relatively stable pricing over time. Crude Oil has a wider spread and multiple peaks, with some prices exceeding $200, indicating periods of significant market fluctuation. Natural Gas is also volatile, with sharp spikes and several extreme outliers near $200. Overall, Crude Oil and Natural Gas experienced greater price variability compared to Coal.

```{r}
# Boxplot of each commodity price with median price
ggplot(energy_long, aes(x = Commodity, y = Price, fill = Commodity)) +
  geom_boxplot() +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = round(..y.., 2)),
    vjust = -0.5,
    color = "black"
  )
  theme_minimal() +
  labs(title = "Boxplot of Energy Commodity Prices (1992–2016)")
```
The boxplot indicates that Coal had the lowest median price at around $50, while Crude Oil and Natural Gas had higher medians near $70. Coal also showed the least variability, with a compact interquartile range and fewer extreme values. In contrast, Crude Oil and Natural Gas had wider spreads, due to more frequent and significant price changes.


## Time Series Analysis of Longitudinal Price Changes in Energy Commodities
To examine long-term trends, we plot each commodity’s price over time. This time series visualization helps us spot major global events or market shifts that caused sudden price changes. It's also useful for identifying patterns or cycles in the data.

```{r vis2}
# Distribution of each commodity price over time
ggplot(energy_long, aes(x = Date, y = Price, color = Commodity)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Energy Commodity Prices Over Time (1992–2016)")
```
The time series plot shows that all three energy commodities generally rose in price from the early 2000s until around 2008, followed by a sharp drop during the global financial crisis. Crude Oil and Natural Gas show strong alignment in trends, with multiple peaks and steep declines, especially around 2008 and 2014. Coal prices were less volatile overall and did not experience changes as sharp as those seen in Crude Oil and Natural Gas, which were more reactive to global market shocks.


## Correlation Analysis
Finally, we calculate the correlation matrix among the three commodities to quantify how closely their prices move together. We use ggcorrplot() to visualize the matrix, with labeled correlation coefficients and color-coded strength.

```{r}
# Correlation matrix
cor_matrix <- cor(energy_data %>% select(-Date), use = "complete.obs")

# Visualize
ggcorrplot(cor_matrix, method = "circle", lab = TRUE, title = "Correlation Between Energy Commodities")
```
The correlation matrix shows strong positive relationships between all three commodities. Crude Oil and Natural Gas have the highest correlation, meaning their prices tend to move closely together. Coal is also strongly correlated with both Crude Oil (0.88) and Natural Gas (0.89), though slightly less. These values suggest that, despite differences in volatility and market behavior, all three energy commodities tend to follow similar overall price trends.


## Conclusion




