---
title: "EDA Code Notebook"
author: "Kimberly Cardinale, Noelle Crichton, and William Kong"
output: html_document
date: "2025-07-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction:
This dataset contains monthly price data for a wide array of global commodities and price indices from January 1980 to June 2016. The data is sourced from the International Monetary Fund’s Primary Commodity Price System and compiled by the dataset’s [GitHub repository](https://github.com/datasets/commodity-prices/tree/main).

Given the significant impact of energy resources on the world economy, this analysis focuses on three major energy commodities: Crude Oil, Natural Gas, and Coal. Analyzing the historical price movements and relationships between these commodities can provide valuable insights into global economic trends, the impacts of energy policies, and how fossil fuels are economically linked.

This exploratory data analysis (EDA) seeks to:  

1. Explore the price distributions for each commodity.  

2. Visualize long-term trends in prices.  

3. Quantify correlations between the commodities.

This process aims to lay the foundations for exploring research questions such as:  

1. How closely are energy commodities correlated?  

2. What global events might explain sharp price fluctuations?


## Loading Packages and Commodity Data
```{r}
library(tidyverse)
library(data.table)
library(lubridate)
library(ggcorrplot)

# Load the dataset
data <- read_csv("https://raw.githubusercontent.com/datasets/commodity-prices/refs/heads/main/data/commodity-prices.csv")
```

```{r load}
data<-read.csv("https://raw.githubusercontent.com/datasets/commodity-prices/refs/heads/main/data/commodity-prices.csv")
```


## Data Wrangling: Date Formatting, Filtering, Renaming, and Missing Value Check
```{r sum}
# Check column names
colnames(data)

# Convert Date to proper format
data <- data %>%
  mutate(Date = ymd(Date))

# Subset only energy-related commodities
energy_data <- data %>%
  filter(Date >= ymd("1992-01-01") & Date <= ymd("2017-06-01")) %>%
  select(Date, `Crude.Oil.petroleum`, `Fuel.Energy.Index`, Coal)

# Rename columns for easier reference
colnames(energy_data) <- c("Date", "CrudeOil", "NaturalGas", "Coal")

# Check for missing values
summary(energy_data)

# Display first 6 rows of cleaned dataset
head(energy_data)
```

The original, raw dataset consisted of 450 rows and 64 columns. Due to missing values in the `NaturalGas` variable from 1980 to 1991, we filtered the data by year to include only the period from 1992 through 2017. The cleaned dataset contains 306 rows and 4 columns, making it ready for exploratory data analysis and visualization.

## Summary Statistics
```{r}
# Summary statistics
energy_data %>%
  select(-Date) %>%
  summarise_all(list(
    mean = ~mean(., na.rm = TRUE),
    median = ~median(., na.rm = TRUE),
    sd = ~sd(., na.rm = TRUE),
    min = ~min(., na.rm = TRUE),
    max = ~max(., na.rm = TRUE)
  ))
```


## Transforming Data to Long Format for Plotting
```{r reshape}
# Reshape to long format for visualization
energy_long <- energy_data %>%
  pivot_longer(cols = -Date, names_to = "Commodity", values_to = "Price")
```


## Visualizing Price Distribution and Median Values by Commodity
```{r vis}
# Histogram of each commodity price
ggplot(energy_long, aes(x = Price)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 40) +
  facet_wrap(~Commodity, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Energy Commodity Prices (1992–2016)")
```


```{r}
# Boxplot of each commodity price with median price
ggplot(energy_long, aes(x = Commodity, y = Price, fill = Commodity)) +
  geom_boxplot() +
  stat_summary(
    fun = median,
    geom = "text",
    aes(label = round(..y.., 2)),
    vjust = -0.5,
    color = "black"
  )
  theme_minimal() +
  labs(title = "Boxplot of Energy Commodity Prices (1992–2016)")
```


## Time Series Analysis of Longitudinal Price Changes in Energy Commodities
```{r vis2}
# Distribution of each commodity price over time
ggplot(energy_long, aes(x = Date, y = Price, color = Commodity)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Energy Commodity Prices Over Time (1992–2016)")
```


## Correlation Analysis
```{r}
# Correlation matrix
cor_matrix <- cor(energy_data %>% select(-Date), use = "complete.obs")

# Visualize
ggcorrplot(cor_matrix, method = "circle", lab = TRUE, title = "Correlation Between Energy Commodities")
```


## Conclusion




