---
title: "EDA Sample"
output: html_document
date: "2025-07-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Introduction: 
This dataset contains monthly price data for global commodities and price indices from January 1980 to December 2016, sourced from the IMF Primary Commodity Price System and compiled by the datasets GitHub repository.

Source: GitHub  https://raw.githubusercontent.com/datasets/commodity-prices/refs/heads/main/data/commodity-prices.csv

Time Coverage: 1980–2016 (monthly)

Size: ~444 rows × 66 columns

We’ll focus our EDA on energy-related commodities: Crude Oil, Natural Gas, and Coal.



```{r}
library(tidyverse)
library(data.table)
library(lubridate)
library(ggcorrplot)

# Load the dataset
df <- "https://raw.githubusercontent.com/datasets/commodity-prices/refs/heads/main/data/commodity-prices.csv"
data <- read_csv(df)
```



```{r load}
data<-read.csv("https://raw.githubusercontent.com/datasets/commodity-prices/refs/heads/main/data/commodity-prices.csv")
```

summary statistics
```{r sum}
# Check column names
colnames(data)

# Convert Date to proper format
data <- data %>%
  mutate(Date = ymd(Date))

# Subset only energy-related commodities
energy_data <- data %>%
  select(Date, `Crude.Oil.petroleum`, `Fuel.Energy.Index`, Coal)

# Rename columns for easier reference
colnames(energy_data) <- c("Date", "CrudeOil", "NaturalGas", "Coal")
```


```{r reshape}
# Check for missing values
summary(energy_data)

# Reshape to long format for visualization
energy_long <- energy_data %>%
  pivot_longer(cols = -Date, names_to = "Commodity", values_to = "Price")
```

```{r vis}
# Histogram of each commodity price
ggplot(energy_long, aes(x = Price)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 40) +
  facet_wrap(~Commodity, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Energy Commodity Prices (1980–2016)")
```

```{r}
# Summary statistics
energy_data %>%
  select(-Date) %>%
  summarise_all(list(
    mean = ~mean(., na.rm = TRUE),
    median = ~median(., na.rm = TRUE),
    sd = ~sd(., na.rm = TRUE),
    min = ~min(., na.rm = TRUE),
    max = ~max(., na.rm = TRUE)
  ))
```

```{r}
# Correlation matrix
cor_matrix <- cor(energy_data %>% select(-Date), use = "complete.obs")

# Visualize
ggcorrplot(cor_matrix, method = "circle", lab = TRUE, title = "Correlation Between Energy Commodities")
```


