---
title: "EDA Code Notebook"
author: "Kimberly Cardinale, Noelle Crichton, and William Kong"
date: "2025-07-17"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction:

Research Motive: analyze the trend of natural resources index within commodity and PPI -- therefore investigate the effect of the former on latter or the dynamic between these two indexes
  Change in commodity Index often preludes the change in PPI -- because commodity prices are upstream inputs in the production process, and PPI reflects the cost of finished goods after those inputs have been processed
    Commodity Index: def
    PPI: def
   
The dynamic can provide valuable insights into global economic trends, the impacts of energy policies on the demand side. 
  
Data Provenance: 
  context -- "data bite man" -- what info do readers need to understand the dataset
  what is the data originally collected for (larger survey conducted every year? government consensus report?) -- what's the purpose and how it aligns with our purposes
  biases in the dataset, constraints
  
  


## Loading Packages and Commodity Data
```{r}
library(tidyverse)
library(data.table)
library(lubridate)
library(ggcorrplot)
library(quantmod)
library(reshape2)
```

```{r load}
Commodity<-read.csv("https://raw.githubusercontent.com/datasets/commodity-prices/refs/heads/main/data/commodity-prices.csv")
PPI<-read.csv("https://raw.githubusercontent.com/kimcardinale/qtm302w-eda-project/d7ba7b915d24af7180918f55ce5ebb5961929ec7/PPIACO.csv")
```

## Data Wrangling: Date Formatting, Filtering, Renaming, and Missing Value Check
This step ensures the dataset is clean and focused on the relevant commodities. We begin by checking the column names and then convert the Date column into proper date format using ymd() from lubridate.

We filter the dataset to include only observations from 1992 to 2017, since earlier years contain missing values for Natural Gas. From the 64-column dataset, we isolate three key variables: Crude Oil, Natural Gas (stored in Fuel.Energy.Index), and Coal. The selected columns are renamed for easier reference, and we confirm the dataset’s structure using summary() and head().

```{r sum}
# Convert Date to proper format and extract Year-Month string
Commodity <- Commodity %>%
  mutate(Date = ymd(Date))

# Subset only energy-related commodities
energy_data <- Commodity %>%
  filter(Date >= ymd("1992-01-01") & Date <= ymd("2016-06-30")) %>%
  select(Date, `Crude.Oil.petroleum`, `Fuel.Energy.Index`, Coal)

# Rename columns
colnames(energy_data) <- c("Date", "CrudeOil", "NaturalGas", "Coal")

# Preview
head(energy_data)
colnames(PPI)
```

``` {r ppi}
PPI <- PPI %>%
  rename(Date = observation_date) %>%     # rename the column
  mutate(Date = ymd(Date)) %>%            # convert to Date format
  filter(Date >= ymd("1992-01-01") & Date <= ymd("2016-06-30"))

merged_data <- inner_join(energy_data, PPI, by = "Date") 
# Preview merged data
head(merged_data)
```


## 4. Summarize and Reshape Data

```{r 3}
summary(merged_data)
head(merged_data)
```

## 5. Visualize Distributions

## Transforming Data to Long Format for Plotting
To facilitate visualization using ggplot2, we reshape the data from wide to long format. This step is important for plotting multiple commodities in a single figure using facet_wrap() or color grouping.

```{r reshape}
# Reshape to long format for visualization
all_long <- merged_data %>%
  select(Date, CrudeOil, NaturalGas, Coal, PPIACO) %>%
  pivot_longer(cols = -Date, names_to = "Index", values_to = "Price")
str(all_long)
```
## Time Series Analysis of Longitudinal Price Changes in Energy Commodities
To examine long-term trends, we plot each commodity’s price over time. This time series visualization helps us spot major global events or market shifts that caused sudden price changes. It's also useful for identifying patterns or cycles in the data.
```{r basic}
# Distribution of each commodity price over time
ggplot(all_long, aes(x = Date, y = Price, color = Index)) +
  geom_line() +
  theme_minimal() +
  labs(
    title = "Energy Commodities and PPI Over Time (1992–2016)",
    x = "Date",
    y = "Price / Index Value",
    color = "Asset"
  )
```


```{r}
merged_data_long <- merged_data %>%
  pivot_longer(cols = c(CrudeOil, NaturalGas, Coal, PPIACO), names_to = "Asset", values_to = "Price")

ggplot(merged_data_long, aes(x = Price)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~Asset, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Commodity and Equity Prices", x = "Price", y = "Frequency")
```

## 6. Descriptive Statistics

```{r 2}
merged_data %>%
  select(CrudeOil, NaturalGas, Coal, PPIACO) %>%
  summarise_all(list(mean = mean, median = median, sd = sd), na.rm = TRUE)
```

## 7. Correlation Analysis

```{r 5}
# Calculate correlation matrix
cor_matrix <- merged_data %>%
  select(CrudeOil, NaturalGas, Coal, PPIACO) %>%
  cor(use = "complete.obs")
cor_melted <- melt(cor_matrix)

ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
  theme_minimal() +
  labs(title = "Correlation Heatmap", x = "", y = "")
```

Lagged Correlation Analysis
```{r lagged cor}
lagged_df <- merged_data %>%
  select(Date, CrudeOil, NaturalGas, Coal, PPIACO) %>%
  mutate(
    CrudeOil_lag1 = lag(CrudeOil, 1),
    CrudeOil_lead1 = lead(CrudeOil, 1),
    
    NaturalGas_lag1 = lag(NaturalGas, 1),
    NaturalGas_lead1 = lead(NaturalGas, 1),
    
    Coal_lag1 = lag(Coal, 1),
    Coal_lead1 = lead(Coal, 1)
  )

cor_results <- tibble(
  Commodity = c("Crude Oil", "Crude Oil", "Crude Oil",
                "Natural Gas", "Natural Gas", "Natural Gas",
                "Coal", "Coal", "Coal"),
  Lag = c("-1 month", "0 month", "+1 month",
          "-1 month", "0 month", "+1 month",
          "-1 month", "0 month", "+1 month"),
  Correlation = c(
    cor(lagged_df$CrudeOil_lag1, lagged_df$PPIACO, use = "complete.obs"),
    cor(lagged_df$CrudeOil, lagged_df$PPIACO, use = "complete.obs"),
    cor(lagged_df$CrudeOil_lead1, lagged_df$PPIACO, use = "complete.obs"),
    
    cor(lagged_df$NaturalGas_lag1, lagged_df$PPIACO, use = "complete.obs"),
    cor(lagged_df$NaturalGas, lagged_df$PPIACO, use = "complete.obs"),
    cor(lagged_df$NaturalGas_lead1, lagged_df$PPIACO, use = "complete.obs"),
    
    cor(lagged_df$Coal_lag1, lagged_df$PPIACO, use = "complete.obs"),
    cor(lagged_df$Coal, lagged_df$PPIACO, use = "complete.obs"),
    cor(lagged_df$Coal_lead1, lagged_df$PPIACO, use = "complete.obs")
  )
)

library(knitr)
kable(cor_results, digits = 3, caption = "Lagged Correlation Between Energy Commodities and PPIACO")

```

rolling correlation analysis
```{r lag effect}
roll_cor <- function(x, y, window = 12) {
  rollapplyr(1:length(x), width = window, FUN = function(i) {
    cor(x[i:(i + window - 1)], y[i:(i + window - 1)], use = "complete.obs")
  }, by = 1, fill = NA)
}

merged_data <- merged_data %>%
  mutate(roll_cor_crude = roll_cor(CrudeOil, PPIACO, 12),
         roll_cor_gas = roll_cor(NaturalGas, PPIACO, 12),
         roll_cor_coal = roll_cor(Coal, PPIACO, 12))


# Reshape rolling correlations into long format
roll_long <- merged_data %>%
  select(Date, roll_cor_crude, roll_cor_gas, roll_cor_coal) %>%
  rename(CrudeOil = roll_cor_crude,
         NaturalGas = roll_cor_gas,
         Coal = roll_cor_coal) %>%
  pivot_longer(cols = -Date, names_to = "Commodity", values_to = "RollingCorrelation")


#Plot
ggplot(roll_long, aes(x = Date, y = RollingCorrelation, color = Commodity)) +
  geom_line(size = 0.5) +
  labs(
    title = "12-Month Rolling Correlation with PPIACO (1992–2016)",
    x = "Date",
    y = "Rolling Correlation",
    color = "Commodity"
  ) +
  theme_minimal()
```
crude oil -- mostly strong positive
natural gas -- second strong positive
coal -- weaker/volatile

## Conclusion

This exploratory analysis investigated the relationship between energy-related commodity prices—Crude Oil, Natural Gas, and Coal—and the Producer Price Index (PPI) from 1992 to 2016. Our motivation stemmed from the economic intuition that changes in commodity prices, which represent upstream production inputs, often precede changes in PPI, which reflects output prices at the producer level.

After wrangling and aligning datasets from the IMF and U.S. Bureau of Labor Statistics, we performed a series of descriptive and correlation analyses. The correlation heatmap revealed that Crude Oil and Coal show relatively strong positive correlations with PPI, while Natural Gas shows a weaker but still positive relationship.

More importantly, our cross-correlation and rolling correlation analyses** uncovered temporal dynamics: Crude Oil and Natural gas exhibit leading relationships, with correlation peaks occurring when these commodity indices are lagged by 1–2 months ahead of PPI. This supports the hypothesis that commodity price movements prelude inflationary changes in producer-level output.

These findings have broader implications for economic forecasting and inflation monitoring. Analysts, policymakers, and investors may consider energy commodity trends as early signals of shifts in inflationary pressure at the production level.

Further research could incorporate other upstream indexes (e.g., metals, agricultural inputs), use monthly % changes or log returns, or explore predictive modeling frameworks based on these observed lead-lag structures.




